---
title: "Water quality analysis of River Thames"
author: "S/16/499"
date: "2/28/2021"
classoption: a4paper
output:
  
  
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,comment = NA)
```

```{r libs, include = FALSE}
library(tidyverse)
library(lubridate)
library(data.table)
library(pastecs)
library(tidyr)
library(factoextra)
library(NbClust)
library(rgl)
```

# Water Quality of River Thames


### Importing the dataset
```{r}
waterQualitydf<-read.csv("River_Thames_Water_Quality.csv",header = TRUE)
```

### Head of the data
```{r}
head(waterQualitydf)
```

### Changing the data type of last column("Dissolved boron (µg l-1)")
```{r}
waterQualitydf$Ammonium..mg.l.1.NH4.<-as.numeric(waterQualitydf$Ammonium..mg.l.1.NH4.)
waterQualitydf$Dissolved.silicon..mg.l.1.Si.<-as.numeric(waterQualitydf$Dissolved.silicon..mg.l.1.Si.)
waterQualitydf$Dissolved.fluoride..mg.l.1.<-as.numeric(waterQualitydf$Dissolved.fluoride..mg.l.1.)
waterQualitydf$Dissolved.boron....µg.l.1.<-as.numeric(waterQualitydf$Dissolved.boron....µg.l.1.)
waterQualitydf$Sampling.date..dd.mm.yyyy.<-as.Date(waterQualitydf$Sampling.date..dd.mm.yyyy.,format="%d/%m/%Y")
```

### Counting missing values and removing them
```{r}
waterQualitydf %>%
  select(everything()) %>%  # replace to your needs
  summarise_all(~(sum(is.na(.))))
```

Removing missing values from the data set
```{r}
df_valid<-na.omit(waterQualitydf)
head(df_valid)
```

Re-check
```{r}
df_valid %>%
  select(everything()) %>%  # replace to your needs
  summarise_all(~(sum(is.na(.))))
```

Types of columns in the dataframe
```{r}
glimpse(df_valid)
```

### Summary of the data set

```{r}
summary(df_valid[-c(1,2,3)])

options(scipen=100)
options(digits=2)
summary_df<-stat.desc(df_valid[-c(1,2,3)])

write.csv(summary_df,"summary.csv",row.names = TRUE)
```

### Plots and diagrams drawn versus time(Month)

```{r}

#unique(df_valid$Site)

df_newBridge<-df_valid %>%
  select(everything()) %>%
  filter(Site=="River Thames at Newbridge")
#head(df_newBridge) 

ggplot(df_newBridge)+geom_line(aes(x=Sampling.date..dd.mm.yyyy.,y=Water.temperature...C.),color="red2",size=1)+
  labs(x="Sampling date",y="Water temperature",title = "River Thames at Newbridge",subtitle = "Water temperature vs Sampling date")
                      
ggplot(df_newBridge)+geom_line(aes(x=Sampling.date..dd.mm.yyyy.,y=Dissolved.boron....µg.l.1.),color="red2",size=1)+
  labs(x="Sampling date",y="Dissolved Boron",title = "River Thames at Newbridge",subtitle = "Dissolved Boron vs Sampling date")

ggplot(df_newBridge)+geom_line(aes(x=Sampling.date..dd.mm.yyyy.,y=Dissolved.sodium..mg.l.1.),color="red2",size=1)+
  labs(x="Sampling date",y="Dissolved Sodium",title = "River Thames at Newbridge",subtitle = "Dissolved Sodium vs Sampling date")

ggplot(df_newBridge)+geom_line(aes(x=Sampling.date..dd.mm.yyyy.,y=Dissolved.nitrate......mg.l.1.NO3.),color="red2",size=1)+
  labs(x="Sampling date",y="Dissolved Nitrate",title = "River Thames at Newbridge",subtitle = "Dissolved Nitrate vs Sampling date")
```

```{r}
ggplot(df_newBridge)+geom_point(aes(x=Dissolved.sodium..mg.l.1.,y=Dissolved.nitrate......mg.l.1.NO3.),color="red2",size=1)+
  labs(x="Dissolved Sodium",y="Dissolved Nitrate",title = "River Thames at Newbridge",subtitle = "Dissolved Nitrate vs Dissolved Sodium")
```

## Manova
MANOVA analysis was done at 5% significance level
    H0 : Mean water properties of each month is equal
                vs
    H1: Mean water properties of at least 2 months are not equal

Creating summary_df new data frame to do MANOVA
```{r}
df_manova<-df_valid %>% 
  select(everything()) %>%
  group_by(Site)

samplingYM<-format(df_manova$Sampling.date..dd.mm.yyyy.,"%Y-%m")

df_manova$samplingYM<-samplingYM

df_manova= subset(df_manova, select = -c(Sampling.time..hh.mm.,Sampling.date..dd.mm.yyyy.) )
head(df_manova)
```

Mean values when grouped by Site and Sample taken date

dependent variable extraction
```{r}
#d.v =dependent variables
d.v<-as.matrix(df_manova
               [2:18])
```

***By SITE***
```{r}
df_groupedbySite<-aggregate(d.v~df_manova$Site,data = df_manova, function(x)round(mean(x),2))
colnames(df_groupedbySite)[1]<-"Site"
head(df_groupedbySite,n=10L)
```
***By Sample taken date***
```{r}
head(aggregate(d.v~df_manova$samplingYM,data = df_manova,function(x)round(mean(x),2)),n=10L)
```

### Manova Test
```{r}
waterqualitymodel<-manova(d.v~df_manova$samplingYM*df_manova$Site)

summary(waterqualitymodel,test = "Pillai")
summary(waterqualitymodel,test = "Wilk")
summary(waterqualitymodel,test = "Roy")
summary(waterqualitymodel,test = "Hotelling-Lawley")
```

Manova second method
```{r}
# dv2<-as.matrix(df_manova[2:18])
# dim(dv2)
# output = lm(dv2~samplingYM*Site, data=df_manova,contrasts = list(samplingYM=contr.sum,Site=contr.sum))
# 
# manova_out<-Manova(output,type="II")
# 
# summary(output,multivariate=TRUE)
```


## Clustering

### Estimating the optimal number of clusters

Creating summary_df new data frame for clustering
```{r}
df_forClustering<-aggregate(d.v~df_manova$Site,data = df_manova, function(x)round(mean(x),2))[,-1]
rownames(df_forClustering)<-aggregate(d.v~df_manova$Site,data = df_manova, function(x)round(mean(x),2))[,1]

head(df_forClustering)
```

Scaling data frame (standardizing the data to make variables comparable)
```{r}
df_scaled<-scale(df_forClustering)
```

Determining the optimal number of clusters for k-means clustering by ***Elbow method***
```{r}
fviz_nbclust(df_scaled, kmeans, method = "wss") +
    geom_vline(xintercept = 3, linetype = 2)+
  labs(subtitle = "Elbow method")
```

Therefore we select k=3 as the number of clusters

### Clustering using K-means
```{r}
set.seed(123)

km.res <- kmeans(df_scaled, 3,nstart = 25)
```

Details
```{r}
print(km.res)
```

```{r}
km.res$centers
```


Visualizing the clusters
```{r}
fviz_cluster(km.res, data = df_scaled,
             palette = c("#2E9FDF", "#00AFBB", "#ed0000"), 
             geom = "point",
             ellipse.type = "convex"
             )+scale_colour_manual(values = c("darkblue", "darkgreen", "red"))

pc<-princomp(df_scaled)
plot3d(pc$scores[,1:3],col = km.res$cluster,size = 20)
```


### cluster summarising

Attaching clusters to each observation accordingly
```{r}
df_forClustering<-cbind(df_forClustering,cluster=km.res$cluster)
```

```{r}
setDT(df_forClustering,keep.rownames = "Site")
head(df_forClustering)
```

```{r}
df_forClustering %>%
  select(everything()) %>%
  filter(cluster==1) 
```

```{r}
mean_clusters<-df_forClustering %>%
  select(-Site) %>%
  group_by(cluster) %>%
  summarise_all("mean")

write.csv(mean_clusters,"meanclusters.csv", row.names = FALSE)
```

```{r}
df_forClustering %>%
  select(Site,cluster) %>%
  group_by(cluster)
```

```{r}
mean_clusters
```

